---
title       : dplyr
subtitle    : with nycflights13 
author      : Most of the tutorial from Kevin Markham
job         : And Hadley Wikham's dplyr resources without which nothing exists
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## What is dplyr?

* Tools for data exploration and transformation
* Intuitive to write and easy to read
* Super-fast on data frames

---
## Installing

```{r eval=FALSE}
if (packageVersion("devtools") < 1.6) {
  install.packages("devtools")
}
devtools::install_github("hadley/lazyeval")
devtools::install_github("hadley/dplyr")
install.packages("nycflights13")
```

---

##  Load our data package

```{r}
library(nycflights13)
suppressMessages(library(dplyr))
```

Package `nycflights13` has data about all flights that departed NYC in 2013 -  5 datasets

1. `flights`:  Flights data
2. `airlines`: Airline names
3. `airports`: Airport metadata
4. `planes` : Plane metadata
5. `weather`: Hourly weather data

---
## Look at `flights` data

```{r}
# explore data
data(flights)
dim(flights)
head(flights,3)
```

336,776 rows... because of its size, this particular dataset has been wrapped, by default, using `tbl_df` (what's that?)

---
## tbl_df

Prints nicely and prevents an accidental display of the whole dataset

```{r}

tblflights <- tbl_df(flights)
head(tblflights,3) # Can also use print(flights,3) instead
```

---
## But if one insists, we can always revert back to a regular dataframe
```{r}
# convert to a normal data frame to see all of the columns
head(data.frame(tblflights),3)
```

---
## Basic single table (df) verbs

1. `filter`: for subsetting variables
2. `select`: for subsetting rows
3. `arrange`: for re-ordering rows
4. `mutate`: for adding new columns
5. `summarise` or `summarize`: for reducing each group to a smaller number of summary statistics 

---

## filter: Keep rows matching criteria

```{r}
# base R: tblflights[tblflights$carrier=="AA" & tblflights$origin=="LGA", ]
filter(tblflights, carrier=="AA" & origin=="LGA")
# same as filter(tblflights, carrier=="AA", origin=="LGA")
```

---
## filter again

```{r}
filter(tblflights, carrier=="AA" | carrier=="UA") 
```

---
## filter again

```{r}
filter(tblflights, carrier %in% c("AA", "UA"))
```

---
## select: Pick columns by name

```{r}
# base R approach to select DepTime, ArrTime, and FlightNum columns
head(tblflights[, c("dep_time", "arr_time", "flight")])
```

---
## select again

```{r}
# dplyr approach
print(select(tblflights, dep_time, arr_time, flight),n=6)
```

---
## select again

```{r}
# use colon to select multiple contiguous columns, and use `contains` to match columns by name
# note: `starts_with`, `ends_with`, and `matches` (for regular expressions) can also be used to match columns by name
head(select(tblflights, flight:dest, contains("arr"), contains("dep")))
```

---
## Chaining over Nesting?

```{r}
# nesting method to select carrier and dep_delay columns and filter for delays over 60 minutes
head(filter(select(tblflights, carrier, dep_delay), dep_delay > 60))
```

---
## Chaining over Nesting

```{r}
tblflights %>%
    select(carrier, dep_delay) %>%
    filter(dep_delay > 60) %>%
    head()
```

---
## Chaining

* Chaining increases readability significantly when there are many commands
* Operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package
* Can be used to replace nesting in R commands outside of dplyr

---
## Chaining

```{r}
# create two vectors and calculate Euclidian distance between them
x1 <- 1:5; x2 <- 2:6

# Usual 
sqrt(sum((x1-x2)^2))

# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```

---
## arrange: Reorder rows

```{r}
# base R approach to select carrier and dep_delay columns and sort by dep_delay
head(tblflights[order(tblflights$dep_delay), c("carrier", "dep_delay")])
```

---
## arrange

```{r}
# dplyr approach
tblflights %>%
    select(carrier, dep_delay) %>%
    arrange(dep_delay) %>% # arrange(desc(dep_delay)) for descending order
    head()
```

---
## mutate: create new variables that are functions of existing variables

```{r}
# base R approach to create a new variable - sum of squares of delays (arr and dep)
tblflights$delaysquare <- tblflights$dep_delay^2 + tblflights$arr_delay^2
head(tblflights[, c("dep_delay", "arr_delay", "delaysquare")])

```

---
## mutate

```{r}
# dplyr approach (prints the new variable but does not store it)
tblflights %>%
    select(dep_delay, arr_delay) %>%
    mutate(delaysquare = dep_delay^2 + arr_delay^2) %>%
    head()

# store the new variable: tblflights <- tblflights %>% 
#         mutate(delaysquare = dep_delay^2 + arr_delay^2) 
```

---
## summarise/summarize: Reduce multiple variables to values

* Primarily useful with data that has been grouped by one or more variables
* `group_by` creates the groups that will be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r}
# base R approaches to calculate the mean arrival delays at different airports
aggregate(arr_delay ~ origin, tblflights, mean)
# or with(tblflights, tapply(arr_delay, origin, mean, na.rm=TRUE))
```

---
## summarise

```{r}
# dplyr approach: create a table grouped by origin, and then summarise each group by taking the mean of arr_delay
tblflights %>%
    group_by(origin) %>%
    summarise(avg_delay = mean(arr_delay, na.rm=TRUE))
```

---
## summarise_each/mutate_each: apply the same summary/mutate function(s) to multiple columns at once

```{r}
# for each carrier, calculate the mean arrival and departure delays at the different origin airports
tblflights %>%
    group_by(origin) %>%
    summarise_each(funs(mean(.,na.rm=TRUE)), arr_delay, dep_delay) %>%
    head()
```

---
## summarise_each

```{r}
# for each carrier, calculate the minimum and maximum of arrival and departure delays
tblflights %>%
    group_by(carrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("_delay")) %>%
    head()
```

---
## mutate_each 
```{r}

tblflights %>%
    select(matches("_delay")) %>%
    head(3)

tblflights %>%
    select(matches("_delay")) %>%
    mutate_each(funs(half=./2)) %>%
    head(3)
```

---
## n():  counts the number of rows in a group

```{r}
# for each month, count the total number of flights each day and sort in order of busiest days
tblflights %>%
    group_by(month,day) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count)) %>%
    head()
```

```{r eval=FALSE}
# rewrite more simply with the `tally` function
tblflights %>%
    group_by(month, day) %>%
    tally(sort = TRUE)
```

---
## n_distinct(vector): counts the number of unique items in that vector

```{r}
# for each destination, count the total number of flights and the number of distinct planes that flew there
tblflights %>%
    group_by(dest) %>%
    summarise(flight_count = n(), plane_count = n_distinct(tailnum)) %>%
    head()
```

---
## Grouping without summarising

```{r}
# for each destination, show the number of flights from the 3 origin airports 
tblflights %>%
    group_by(dest) %>%
    select(origin) %>%
    table() %>%
    head()
```

---
## Window Functions

* Aggregation function (like `mean`) takes n inputs and returns 1 value
* [Window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) takes n inputs and returns n values
* Includes ranking and ordering functions (like `min_rank`), offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`).

---
## Try this

```{r}
# for each carrier, calculate which two days of the year they had their longest departure delays
# note: smallest (not largest) value is ranked as 1, so you have to use `desc` to rank by largest value
tblflights %>%
    group_by(carrier) %>%
    select(month, day, dep_delay) %>%
    filter(min_rank(desc(dep_delay)) <= 2 & dep_delay!="NA") %>%
    arrange(carrier, desc(dep_delay)) %>%
    head()
```

---
## Other things to play with

```{r eval=FALSE}
# for each carrier, calculate which two days of the year they had their longest departure delays--- rewrite previous with the `top_n` function
tblflights %>%
    group_by(carrier) %>%
    select(month, day, dep_delay) %>%
    filter(dep_delay!="NA") %>%
    top_n(2) %>%
    arrange(carrier, desc(dep_delay)) %>% head()

# for each month, calculate the number of flights and the change from the previous month
tblflights %>%
    group_by(month) %>%
    summarise(flight_count = n()) %>%
    mutate(change = flight_count - lag(flight_count))

# rewrite previous with the `tally` function
tblflights %>%
    group_by(month) %>%
    tally() %>%
    mutate(change = n - lag(n))
```

---
## Other Useful Convenience Functions

```{r eval=FALSE}
# randomly sample a fixed number of rows, without replacement
tblflights %>% sample_n(5)

# randomly sample a fraction of rows, with replacement
tblflights %>% sample_frac(0.25, replace=TRUE)

# base R approach to view the structure of an object
str(tblflights)

# dplyr approach: better formatting, and adapts to your screen width
glimpse(tblflights)
```

---
## do : for doing arbitrary operations

```{r}
model=tblflights %>% group_by(origin) %>% do(lm=lm(dep_delay~arr_delay+carrier,data=.)) 
model %>% summarise(rsq=summary(lm)$r.squared)
```

---
## Binary verbs

4 joins from SQL

* `inner_join(x, y)`: matching x + y
* `left_join(x, y)` : all x + matching y
* `semi_join(x, y)` : all x with match in y
* `anti_join(x, y)` : all x without match in y

---
## Quick look at 3 of 5 nycflights13 datasets
```{r}
tblflights %>% head(1) 
airlines %>% head(1)
airports %>% head(1)
```

---
## inner_join(x,y): matching x + y
```{r}
tblflights %>% inner_join(airlines) %>% head(1)
tblflights %>% inner_join(airlines) %>% select(distance:name) %>% head(1)
```

---
## left_join(x,y): all x + matching y

```{r}
faatblflights=tblflights %>% select(origin) %>% mutate(faa=origin)
faatblflights %>% left_join(airports) %>% head(2) 
airports %>% left_join(faatblflights) %>% head(2)
```

---
## semi_join(x, y) : all x with match in y

```{r}
airports %>% left_join(faatblflights) %>% head(2)
airports %>% semi_join(faatblflights) %>% head(2)
```

---

## anti_join(x, y) : all x without match in y
```{r}
faatblflights %>% anti_join(airports)
airports %>% anti_join(faatblflights)
```

---
## The fun has to stop here...

* [Source Code of this slidify presentation](https://github.com/patilv/dplyr-nycflights)
* [Kevin Markham's tutorial](http://www.dataschool.io/dplyr-tutorial-for-faster-data-manipulation-in-r/)
* [Official dplyr reference manual and vignettes on CRAN](http://cran.r-project.org/web/packages/dplyr/index.html): vignettes are well-written and cover many aspects of dplyr
* [July 2014 webinar about dplyr (and ggvis) by Hadley Wickham](http://pages.rstudio.net/Webinar-Series-Recording-Essential-Tools-for-R.html) and related [slides/code](https://github.com/rstudio/webinars/tree/master/2014-01): mostly conceptual, with a bit of code
* [dplyr tutorial by Hadley Wickham](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) at the [useR! 2014 conference](http://user2014.stat.ucla.edu/): excellent, in-depth tutorial with lots of example code (Dropbox link includes slides, code files, and data files)
* [dplyr GitHub repo](https://github.com/hadley/dplyr) and [list of releases](https://github.com/hadley/dplyr/releases)
